---
title: "Welcome the world of Predictive Ecology"
author: "Tati Micheletti"
date: "24 June 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# INSTRUCTIONS FOR THE PREDICTIVE ECOLOGY COURSE

## 1. PAPERS  
  
For the first part of the course, I would like you to divide the group in two, and each group should chose one of the following two papers to read:

**PERFICT: A Re-imagined foundation for predictive ecology**  
(https://onlinelibrary.wiley.com/doi/full/10.1111/ele.13994)  

  OR  

**Empowering ecological modellers with a PERFICT workflow: Seamlessly linking data, parameterisation, prediction, validation and visualisation**  
(https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.14034)

Please take notes on what you find most important about the papers and we will discuss it in class.  


Now you will find the instructions necessary to navigate the 
practical part of the course on Predictive Ecology. 

## 2. GROUPS  

First, you will need to divide the whole class into 3 groups. 
These will be known at first as the TREE groups, the BIRDS group 
and the TURTLE group. Each group will have to bring one (AND ONLY 
ONE) computer to the course.  

## 3. NEEDED SOFTWARE  
  
The 3 computers should be preferably Windows based. They will have 
to have the following installed:
  2.1. Excel (preferably) or OpenOffice Calc.
  2.2. R
  2.3. RStudio (Desktop, not Server)  

To install the last two, please use the follownig useful illustrated guide: 
https://techvidvan.com/tutorials/install-r/#:~:text=Step%20%E2%80%93%201%3A%20Go%20to%20CRAN,the%20latest%20version%20of%20R.

Alternatively, watch this YouTube Video: https://www.youtube.com/watch?v=9-RrkJQQYqY
with the accompanying link: https://rpubs.com/MohammadShadan/installrrstudio  

Or you can even skip all that, and download the software directly here!  
R: https://cran.r-project.org/bin/windows/base/R-4.1.3-win.exe  
RStudio: https://download1.rstudio.org/desktop/windows/RStudio-2022.02.1-461.exe  
Download and click on the .exe and follow instructions. Use all defaults.  

## 4. INSTALL LIBRARIES  
  
Once you have managed to install the needed software, you should install the 
libraries we will need for the exercise. It is important to do this 
BEFORE the course. Libraries can take time to download and install and we 
don't want to waste time doing it during the course. Ideally, install each new library 
(in other words, run each line below) after restarting your R session (shortcut is Ctrl + Shift + F10). 
This will help prevent errors (which can anyway still happen!).  

Once you have installed R, you can open it and run the code between the lines below,
one at a time.  

```{r install}
if (!require("Require")) {install.packages("Require"); require("Require")}
Require::Require("sf")
Require::Require("googledrive")
Require::Require("RCurl")
Require::Require("data.table")
Require::Require("terra")
Require::Require("reproducible")
Require::Require("ggplot2")
Require::Require("gridExtra")
```


## 5. DON'T PANIC    
  
I know it might be scary at the beginning, especially if something doesn't work. 
But we will go very slowly with the hands-on exercise, and I will guide you through it. 
There is no need to panic. Even if you were not successful installing the libraries, 
the important thing is to try! If you have any errors during library installation, 
please send me an email at `tati.micheletti@gmail.com` with 
*a print screen or a copy of the error* and I will help you. 

## 6. OBJECTIVES  

The main objective of the practical part of this course is to answer the following scientific question 
(with made up data, of course):  

  
**Are birds good umbrellas for turtles now and in the future?**
  
  
in other words  
  
**Do we have a positive relationship between bird abundance and turtle habitat suitability?**  


See you soon!  
    
Tati

