---
title: "Integrating Landscape Change with Birds and Turtles"
author: "Tati Micheletti"
date: "4/7/2022"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Integrating our work  

If we arrived here, it means we were successful in forecasting landscape changes, 
simulating turtle habitat suitability and bird abundance! Thank you all for your 
hard work on it. :)

So far, we have only worked with our own groups, so now we will hear a little bit 
about what other groups did.  

Now we will see how we can integrate all this work and answer our burning scientific 
question: are birds good umbrella species for turtles? Of course, we need to remember 
that this example is completely made up, so none of our conclusions here really 
apply to the real world. However, once we are finished you will have had a hands-on 
experience on how we can work collaboratively with predictive ecology and why 
this knowledge is important in whatever area you work on.  

So let's get started!  

As always, our first step is to load the necessary libraries:  

```{r libraries}
if (!require("Require")) {install.packages("Require"); require("Require")}
Require("data.table")
Require("raster")
Require("googledrive")
Require("reproducible")
Require("ggplot2")
Require("gridExtra")
```

The next thing we will do is to `source` all `functions` that will help us integrate 
all the work done by all three groups:  

```{r sourcing}
source("landscapeSimulation.R")
source("turtleSimulation.R")
source("integratingSimulations.R")
```

Now we are ready to start with the integration!

First, we need to get all the data that was collected and generated by each one 
of the groups.

```{r getData}
allData <- getAllData()
```

The **_Landscape Simulation_** group has generated changes in landscape through 
100 years. These include forest growth, wildfire and human disturbance. Here we 
can see the initial landscape, and the landscape after 100 years:

```{r landscapeResults}
initialLandscape <- getInitialLandscape()
finalLandscape <- ratify(allData$landscape)
plot(initialLandscape)
plot(finalLandscape)
```

The **_Species Abundance Simulation_** group developed a model to predict bird 
abundance based on the landscape. The model can be seen here: 

```{r birdsResults}
birdModel <- allData$birdResults
```

At last, the **_Habitat Selection Simulation_** group developed a habitat selection 
model to predict habitat suitability for turtles. Their results can be seen here, 
in a ranking of the different habitats: 

```{r turtleResults}
habitats <- merge(availableHabitats(visible = FALSE), allData$turtleResults)
```

Now we will try to answer our question: are birds good umbrellas for turtles now 
and in the future?

The first step to answer this question is to map both bird abundance now and in 
the future using our generated landscape.

```{r generateBirdMap}
birdMaps <- generateBirdMaps(initialLandscape, finalLandscape, birdModel)
plot(birdMaps, col = heat.colors(6))
```

We then map turtle habitat suitability now and in the future also using our generated 
landscape.

```{r generateTurtleMap}
turtleMaps <- generateTurtleMaps(initialLandscape, finalLandscape, habitats)
plot(turtleMaps, col = heat.colors(6))
```

Now, to answer our question, we will do a correlation analysis between the two 
sets of maps. To do that, we can use the  

```{r correction}
answerToOurQuestion <- areBirdsGoodUmbrellaForTurtle(birdMaps, turtleMaps)
```

So, can we answer our question now? Are birds good umbrellas for turtles now? And 
how does this relationship changes in the future?  

I hope you have enjoyed this exercise and now understands a little bit more about 
the importance of predictive ecology. Please do not hesitate in contacting me if 
you have doubts about the exercise, or the exchange we had during this course, 
and I look forward to seeing you in a not so distant future. :)

Cheers,

Tati

PS.: If you really enjoyed this exercise or want to try the exercises from other 
groups, but don't have any Lego's at hand, don't despair. I have prepared a 
script that you can run each step of the game with made up data, so you can 
understand each cog of this complex machine. You can run the full game from 
scratch running the script named `gameTestingScript.R`. Have fun! 


