---
title: "Landscape Simulation Exercise"
author: "Tati Micheletti"
date: "4/5/2022"
output: pdf_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Simulating Landscape Changes 

The present game-exercise intends to give you a hands-on experience on how ecological 
modelers use data, statistics, and probability to forecast changes in our planet.
To play the game, you were previously divided in three groups:  
1. the **Turtle** group  
2. the **Bird** group, and    
3. the **Tree** group.  

If you are reading this, you belong to the Tree group, which we will rename 
to **_Landscape Simulation_** group. Fancier, right? :)  

You will be responsible for simulating changes in the landscape and, ultimately, 
influencing how the turtles and birds populations will change.

In the folder you downloaded, you received a spreadsheet named `landscape_simulation.xlsx`. 
This sheet brings the aerial view of your landscape and you can see how it changes through time.

### Getting started  

Before you start, you will need to install two libraries to help with the simulations. 
The first library is called `Require`, and is an R package (or library) dedicated 
to improve code reproducibility on the setup side. The second library we will need 
is called `data.table`, which improves efficiency of data manipulation.

```{r libraries}
if (!require("Require")) {install.packages("Require"); require("Require")}
Require("data.table")
Require("raster")
Require("googledrive")
```

The next thing we will do is to `source` all `functions` that will help us simulate 
the movement of the turtle. These functions are described in the file `turtleSimulation.R`, 
which is a typical coding file for the R environment. Functions are the basis for 
programming. They are generally generic commands to explain to the software what 
to do. If you are interested, you can read more about:  

**R and functions**, in this great and simple tutorial from Norm Matloff: <https://github.com/matloff/fasteR>  
and in a cool YouTube video from R Programming <https://www.youtube.com/watch?v=BvKETZ6kr9Q>  
**data.table and efficient data manipulation** in this tutorial:   <https://www.machinelearningplus.com/data-manipulation/datatable-in-r-complete-guide/>   
**RMarkdown** and other awesome packages from Yihui Xie here: <https://bookdown.org/yihui/rmarkdown/> and   <https://bookdown.org/yihui/rmarkdown-cookbook/>  
**RStudio** in a fun and easy way: <https://moderndive.netlify.app/1-getting-started.html>  

The present `landscapeSimulation.Rmd`, 
on the other hand, is a mixed-file that allows for both documentation (normal text) and 
code to be written. When the file is done, this document can be `rendered` into 
a PDF with a really neat format, which you can also see in the folder, as 
`landscapeSimulation.pdf`.

### Functions' Sourcing 

To `source` (or in other words, _read_) the functions we will using, we can use a 
function in R called `source()` (I agree, not too creative, but easy to remember).
To run the line below, you can press the little green arrow on the right side of 
the `code chunck`, where the instructions for R to source the file with functions 
are written.

```{r sourcing}
source("landscapeSimulation.R")
```

If everything goes according to the plan, you should have seen in the `Console` 
below this window the following sentence:  

`All functions were correctly sourced! You are ready to start.`

Now the real fun begins! 

### Possible fates for different landscape types

For each 10 year time step, we will simulate the fate of 
1. Forests
2. Human disturbance

To make it fun and understandable, we will do a very (VERY!) simple exercise on 
simulation, where water bodies are kept constant through time, climate change is 
ignored (although it should NOT in real life), and the processes for all animals 
and landscape a very simplified. After the exercise we will discuss a study 
that brings in considerably more complexity. But one thing at a time.

First, we will discuss how each element can be changed. Then, we will start 
changing them. 

#### Forests  

There are two important things about forests in our game that you need to 
remember:  
1. Forests have one very important property for animals: **age**.  
2. Forests can have only three behaviors: **grow**, **burn** or be 
**replaced by human development**.  
In our game, older forests are less resilient to fire, followed by middle-aged  
forests, and then by young forests. If forests do not burn nor are 
replaced by human development, they will grow. A forest that is 10 years old today 
will be 20 years old in 10 years, the time step we are using for our landscape 
simulations.

#### Human disturbance  

There is only one level of human disturbance, which can be grey or black. 
The size of new developments will always be 1. Human disturbance can only 
**appear** in a *forested area* or in a *burned area*, being more likely in a 
burned area, followed by a *young forest*, *middle-aged forest* and only then 
by *older forest* (in our game, everyone understand and respect the importance 
of forests!).

### Simulating landscape changes  

First of all, we will create a landscape where the changes will happen:  

```{r initialLandscape}
landscape <- getInitialLandscape()
```

Here we can see the legend of our landscape:
1. Old forest (dark green) 
2. Middle-aged forest (middle green)
3. Young forest (light green)
4. Recently burned patch (light yellow)
5. Human disturbed patch (grey or black)
6. Water (light blue)

Once we have our landscape, we will have changes happening in the following order:
1. Forest growth
2. Human development
3. Fire

#### Forest growth  

To know how forest grew during the decade, you can use the function:  

```{r forestGrowth}
simulateForestGrowth()
```

#### Human development  

To know which cells became disturbed by human actions during the decade, you can use the function:  

```{r humanDisturbance}
simulateHumanDisturbance()
```

#### Fire  

To know which cells burned during the decade, you can use the function:  

```{r fire}
simulateFire()
```

#### Checking changes to landscape  

We will then check which cells changed and follow the changes in landscape.  

```{r changes}
changes <- checkChangesInLandscape()
```

#### Implementing changes to landscape  

At last, we will add our changes to the original landscape created here and 
observe how the landscape changed. In other words, we will update our landscape 
with the observed changes.     

```{r updating}
landscape <- updateLandscape(landscape, changes)
```

Congratulations! Now you reached the end of the second time step of the simulation 
(the first step is considered the initial conditions). You can now keep simulating 
forest growth, human disturbance and fire until you reach 10 time steps.  

Once you finished all simulations, our last step is to save the landscape map 
and the send it to me. We will do that with the code below:  

```{r saving}
saveAndSendResults(landscape)
```

If you never used `googledrive` in R, you will probably have some setup to do. But 
don't worry! Follow the instructions in the screen and it should work.  

If you see the message: `Results uploaded!`, you have finished the exercise! 

Once all groups are done, we will integrate all projects and answer our 
research question. :)

### Playing further on  

If you enjoyed the simulations and would like to see how playing with the probabilities 
of human disturbance and fire would change your landscape, you can restart the game by:
1. Deleting both files `landscapeSimulationThroughTime.csv` and `landscapeResults.tif` 
from the `data` folder in the 
`predictiveEcologyInAction` directory. 
2. Re-running the code contained in this `.Rmd` file while modifying the functions for 
simulation of human disturbance and fire disturbance as such: 
`simulateHumanDisturbance(intensity = newLevel)`  
`simulateFire(intensity = newLevel)`
where `newLevel` is a multiplier value to the probabilities of these two changes. 
For example, for a scenario where human disturbance is 3 times more likely
than the default set up, and the fire is only half likely to happen, you can pass 
the following:
`simulateHumanDisturbance(intensity = 3)`  
`simulateFire(intensity = 0.5)`

I hope you had some fun, but most of all, could grasp the general mechanisms 
behind how we simulate landscape changes.  






